---
title: "Assignment 2: Mapping Fire Incidents and FDNY Response Times"
author: Thomas Brambor
date: 2017-03-07
always_allow_html: yes
output: 
  html_document:
    keep_md: true
    code_folding: hide
---

Fires in NYC and FDNY Response
================================

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Dropbox/QMSS/Data Visualization/Beiers_Sophie/hw2")
library(ggmap)
```

## Overview

For this project, I investigate serious incidents requiring the fire department to respond. Using data about the locations of firehouses and fires occurring in New York City, I investigate whether response times to fires differ across the city. Second, I focus on whether the distance from the firehouse effects response times to the scene.

## Data

### Incidents responded to by fire companies

NYC Open Data has data on all [incidents responded to by fire companies](https://data.cityofnewyork.us/Public-Safety/Incidents-Responded-to-by-Fire-Companies/tm6d-hbzd). I have included the variable description file in the exercise folder. The following variables are available:

  - IM_INCIDENT_KEY:	Unique identifier for each incident which serves
  - INCIDENT_TYPE_DESC	The code and description of the incident category type
  - INCIDENT_DATE_TIME	The date and time that the incident was logged into the Computer Aided Dispatch system
  - ARRIVAL_DATE_TIME	The date and time that the first unit arrived on scene
  - UNITS_ONSCENE	Total number of units that arrived on scene
  - LAST_UNIT_CLEARED_DATETIME	The date and time that the incident was completed and the last unit cleared the scene
  - HIGHEST_LEVEL_DESC	The highest alarm level that the incident received
  - TOTAL_INCIDENT_DURATION	The total number of seconds from when then incident was created to when the incident was closed
  - ACTION_TAKEN1_DESC	The code and description of the first action taken
  - ACTION_TAKEN2_DESC	The code and description of the second action taken
  - ACTION_TAKEN3_DESC	The code and description of the third action taken
  - PROPERTY_USE_DESC	The code and description of the type of street or building where the incident took place
  - STREET_HIGHWAY	The name of the street where the incident_took place
  - ZIP_CODE	The postal zip code where the incident took place
  - BOROUGH_DESC	The borough where the incident took place
  - FLOOR	The floor of the building where the incident took place
  - CO_DETECTOR_PRESENT_DESC	Indicator for when a CO detector was present
  - FIRE_ORIGIN_BELOW_GRADE_FLAG	Indicator for when the fire originated below grade
  - STORY_FIRE_ORIGIN_COUNT	Story in which the fire originated
  - FIRE_SPREAD_DESC	How far the fire spread from the object of origin
  - DETECTOR_PRESENCE_DESC	Indicator for when a  detector was present
  - AES_PRESENCE_DESC	Indicator for when an Automatic Extinguishing System is present
  - STANDPIPE_SYS_PRESENT_FLAG	Indicator for when a standpipe was present in the area of origin of a fire

This dataset is only updated annually, and thus far only data from 2013 to 2015 is contained. The following analyses are run using just 2015 data and only the most severe fires (Level 7).

```{r, eval=FALSE}
fire <- read_csv("./data/Incidents_Responded_to_by_Fire_Companies.csv") 
fire$year <- substr(fire$INCIDENT_DATE_TIME, 7, 10)
fire <- fire%>% 
  filter(HIGHEST_LEVEL_DESC == "7 - Signal 7-5") %>%
  filter(year==2015)
```

```{r, eval=FALSE}
# Geocoding data
# Make list of addresses
address <- str_c( str_to_title(fire$STREET_HIGHWAY),
                  "New York, NY",
                  fire$ZIP_CODE,
                  sep=", ")

# Register Google API Key
register_google(key = Sys.getenv("GOOGLE_MAPS_API_KEY"))

# Geocode Addresses
latlong <- geocode(address, output = c("latlon"))

# Merge on
fire$Latitude  <- latlong$lat
fire$Longitude <- latlong$lon

# Save File
write_csv(fire, "severe_incidents.csv")
```

### FDNY Firehouse Listing

NYC Open Data also provides data on the [location of all 218 firehouses in NYC](https://data.cityofnewyork.us/Public-Safety/FDNY-Firehouse-Listing/hc8x-tcnd). Relevant for our analyses are the following variables:

    FacilityName, Borough, Latitude, Longitude

```{r, warning = F, message = F}
# Read in both datasets 
library(tidyverse)
fire <- read_csv("./data/severe_incidents.csv") 
locs <- read_csv("./data/FDNY_Firehouse_Listing.csv") 
locs <- locs %>% 
  select(FacilityName, Borough, Latitude, Longitude)
```
## Visualizations

### 1. Location of Severe Fires
#### a. Leaflet Map: Incidents

Below is a `leaflet` map of the severe fires contained in the file `severe_incidents.csv`. Fires that occured at locations outside the five boroughs of New York City were excluded. The type of incident, date and time of incident, where the fire spread, the number of firemen/women units that came onto scene, whether a detector was present and the total duration of the incident are available as information in interactive popups. 

```{r}
library(leaflet)
library(RColorBrewer)

# popups
fire$INCIDENT_TYPE_DESC <- substring(fire$INCIDENT_TYPE_DESC, 7) # rid of numbers 
fire$FIRE_SPREAD_DESC <- substring(fire$FIRE_SPREAD_DESC, 5)
fire$DETECTOR_PRESENCE_DESC <- substring(fire$DETECTOR_PRESENCE_DESC, 5)
fire$TOTAL_INCIDENT_DURATION <- round(fire$TOTAL_INCIDENT_DURATION/60, 2) # convert to minutes

content <- paste("What:",fire$INCIDENT_TYPE_DESC,"<br/>",
                 "When:",fire$INCIDENT_DATE_TIME,"<br/>",
                 "Location:",fire$FIRE_SPREAD_DESC,"<br/>",
                 "Length of Incident (min):", fire$TOTAL_INCIDENT_DURATION, "<br/>",
                 "Number of Units On Scene:",fire$UNITS_ONSCENE,"<br/>",
                 "Fire Dectector Status:", fire$DETECTOR_PRESENCE_DESC, "<br/>")

fireIcons <- icons(
  iconUrl = "./fire.png",
  iconWidth = 8, iconHeight = 11
  #iconAnchorX = 22, iconAnchorY = 94
)

# map 
leaflet(data = fire, options = leafletOptions(minZoom = 10)) %>% 
  setView(lat = 40.730610, lng = -73.935242, zoom = 10) %>%
  addTiles() %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  addMarkers(lat = ~Latitude, lng = ~Longitude, 
             icon = fireIcons, popup = content)


```

#### b. Heat Map: Incidents

Out of curiosity, I created a heatmap of the same most severe incidents to investigate whether certain areas in NYC had more 2015 fires than others. From the map we can see Manhattan, the Bronx, and some of Brooklyn experienced the most severe incidents while Queens and Long Island didn't experience quite as many. This is likely due to population density differences between these boroughs. 

```{r}
incident_map <- get_map(location = c(lon = -73.935242, lat = 40.730610), maptype = "toner-lite", zoom = 10)
ggmap(incident_map) + 
 geom_density2d(aes(x = Longitude, y = Latitude), 
  data = full_fire, 
  color="blue", 
  size=0.5, 
  bins=10) +
  stat_density2d(aes(x = Longitude, y = Latitude,
    fill = ..level.., 
    alpha = ..level..),
    data = full_fire, 
    geom = 'polygon', 
    bins = 10,
    show.legend = FALSE) +
  scale_fill_gradient2(low = "yellow", mid = "orange", high = "red") +
  scale_alpha(range = c(0.00, 0.5)) +
  labs(x = '', y = '') +
  theme(axis.text = element_blank()) +
  theme(axis.ticks = element_blank())
```

### 2. Layers and Clusters

#### a. Leaflet Map: Property Type

Beginning with the previous leaflet map, I distinguished the markers of the fire locations by `PROPERTY_USE_DESC`, i.e. what kind of property was affected. There were many categories of properties, so I rounded them down as much as possible while still providing interesting information. Information about the property type as well the same details about the incidents are recorded in interactive popups. 

```{r}
##### Data Cleaning
fire$PROPERTY_USE_DESC <- substring(fire$PROPERTY_USE_DESC, 7) # rid of numbers before prop name
tab <- as.data.frame(table(fire$PROPERTY_USE_DESC)) # table to get frequency
colnames(tab) <-  c("PROPERTY_USE_DESC", "Freq")
fireprop <- merge(x = fire, y = tab, by = "PROPERTY_USE_DESC", all = TRUE) # merge df and table 
fireprop$PROPERTY_USE_DESC <- ifelse(fireprop$Freq < 3, "Other", fireprop$PROPERTY_USE_DESC) # all property types with fewer than 10 occurences become "other"

## Renaming and Classifying

# Open spaces
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Open land or field"] <- "Open Land/Field"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Outside or special property, other"] <- "Open Land/Field"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Dock, marina, pier, wharf"] <- "Open Land/Field"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Forest, timberland, woodland"] <- "Open Land/Field"

# schools
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Schools, non-adult, other"] <- "School"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Adult education center, college classroom"] <- "School"

# residential spaces
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Residential, other"] <- "Residential Home"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Multifamily dwelling"] <- "Residential Home"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="1 or 2 family dwelling"] <- "Residential Home"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Dormitory-type residence, other"] <- "Residential Home"

# streets
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Street, other"] <- "Street/Highway"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Residential street, road or residential driveway"] <- "Street/Highway"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Street or road in commercial area"] <- "Street/Highway"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Railroad right-of-way"] <- "Street/Highway"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Highway or divided highway"] <- "Street/Highway"

# parking garages
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Parking garage, (detached residential garage)"] <- "Parking Garage"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Vehicle parking area"] <- "Parking Garage"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Vehicle storage, other"] <- "Parking Garage"

# hotels
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Hotel/motel, commercial"] <- "Hotel/Motel"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Boarding/rooming house, residential hotels"] <- "Hotel/Motel"

# businesses
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Mercantile, business, other"] <- "Mercantile Business"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Laundry, dry cleaning"] <- "Mercantile Business"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Business office"] <- "Business Office"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="General retail, other"] <- "Mercantile Business"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Specialty shop"] <- "Mercantile Business"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Professional supplies, services"] <- "Mercantile Business"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Food and beverage sales, grocery store"] <- "Mercantile Business"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Convenience store"] <- "Mercantile Business"

# restaurants
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Restaurant or cafeteria"] <- "Restaurant/Bar"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Eating, drinking places, other"] <- "Restaurant/Bar"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Bar or nightclub"] <- "Restaurant/Bar"

# warehouses
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Property Use, other"] <- "Warehouse/Storage"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Storage, other"] <- "Warehouse/Storage"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Warehouse"] <- "Warehouse/Storage"

# others
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Hospital - medical or psychiatric"] <- "Hospital"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Motor vehicle or boat sales, services, repair"] <- "Gas Station/Vehicle Repair"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Service station, gas station"] <- "Gas Station/Vehicle Repair"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Rapid transit station"] <- "Transit Station"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Manufacturing, processing"] <- "Other"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Outbuilding or shed"] <- "Other"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Undetermined"] <- "Other"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Industrial plant yard - area"] <- "Other"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Church, mosque, synagogue, temple, chapel"] <- "Other"
fireprop$PROPERTY_USE_DESC[fireprop$PROPERTY_USE_DESC=="Construction site"] <- "Other"
```


```{r}
##### Map
# create order of factor based on how many incidents occured in that type of property
fireprop$PROPERTY_USE_DESC <- factor((fireprop$PROPERTY_USE_DESC), levels = c("Hospital", "Transit Station", "School", "Hotel/Motel", "Gas Station/Vehicle Repair", "Warehouse/Storage", "Parking Garage", "Business Office", "Restaurant/Bar", "Open Land/Field", "Street/Highway", "Other", "Mercantile Business", "Residential Home"))
#colors <- c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "#c6dbef", "#ffffcc", "#ffeda0", "#fed976", #"#feb24c", "#fd8d3c", "#fc4e2a", "#e31a1c", "#bd0026", "#800026")
colorCount <- length(unique(fireprop$PROPERTY_USE_DESC)) # number of levels
colors = colorRampPalette(brewer.pal(9, "YlOrRd"))(colorCount)

# popups

content <- paste("What:",fireprop$INCIDENT_TYPE_DESC,"<br/>",
                 "When:",fireprop$INCIDENT_DATE_TIME,"<br/>",
                 "Where:",fireprop$PROPERTY_USE_DESC,"<br/>", # added where
                 "Location:",fireprop$FIRE_SPREAD_DESC,"<br/>",
                 "Length of Incident (min):", fireprop$TOTAL_INCIDENT_DURATION, "<br/>",
                 "Number of Units On Scene:",fireprop$UNITS_ONSCENE,"<br/>",
                 "Fire Dectector Status:", fireprop$DETECTOR_PRESENCE_DESC, "<br/>")

# map -- starting with other map but changed data source 
leaflet(data = fireprop, options = leafletOptions(minZoom = 9)) %>% 
  setView(lat = 40.730610, lng = -73.935242, zoom = 9) %>%
  addTiles() %>% 
  addProviderTiles("CartoDB.DarkMatter") %>% 
  addCircles(lat = ~Latitude, lng = ~Longitude, 
             color = colors, 
             weight = 5,
             opacity = 0.8,
             fill = TRUE,
             fillOpacity = 0.8,
             popup = content ) %>% 
  addLegend("bottomright", 
            colors = c("#FFFFCC", "#FFF3B0", "#FEE896", "#FEDC7C", "#FEC662", "#FDAF4A", "#FD9840", "#FC7936", "#FC522B", "#EE3222", "#DD151D", "#C50523","#A50026","#800026"), 
            values = ~fireprop$PROPERTY_USE_DESC, 
            title = "Incident Property Type",
            labels = c("Hospital", "Transit Station", "School", "Hotel/Motel", "Gas Station/Vehicle Repair", "Warehouse/Storage", "Parking Garage", "Business Office", "Restaurant/Bar", "Open Land/Field", "Street/Highway", "Other", "Mercantile Business", "Residential Home"))


#colors = c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "#c6dbef", "#ffffcc", "#ffeda0", "#fed976", #"#feb24c", "#fd8d3c", "#fc4e2a", "#e31a1c", "#bd0026", "#800026") # blue/red
```

#### b. Clusters

I added marker clustering, so that zooming in reveals the individual locations but the zoomed out map only shows the clusters.

```{r}
leaflet(data = fireprop, options = leafletOptions(minZoom = 10)) %>% 
  setView(lat = 40.730610, lng = -73.935242, zoom = 10) %>%
  addTiles() %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  addCircleMarkers(lat = ~Latitude, lng = ~Longitude, 
                   color = colors, 
                   popup = content, 
                   clusterOptions = markerClusterOptions()) %>% 
  addLegend("bottomright", 
            colors = c("#FFFFCC", "#FFF3B0", "#FEE896", "#FEDC7C", "#FEC662", "#FDAF4A", "#FD9840", "#FC7936", "#FC522B", "#EE3222", "#DD151D", "#C50523","#A50026","#800026"), 
            values = ~fireprop$PROPERTY_USE_DESC, 
            title = "Incident Property Type",
            labels = c("Hospital", "Transit Station", "School", "Hotel/Motel", "Gas Station/Vehicle Repair", "Warehouse/Storage", "Parking Garage", "Business Office", "Restaurant/Bar", "Open Land/Field", "Street/Highway", "Other", "Mercantile Business", "Residential Home"))

```


### 3. Fire Houses

Next, I adjusted the initial leaflet map such that the size of the circle markers indicate severity (by the number of units that came on to the scene) and colors indicated the length of time the fire persisted. Firehouse locations were also added. There are two layers to the map ("Incidents", "Firehouses") that allow the user to select which information to show. 

```{r}
# colors
fireprop$brks <- cut(fireprop$TOTAL_INCIDENT_DURATION, 
                   breaks=c(0, 75, 110, 120, 150, 7140), 
                   labels=c("10-75 min", "1.25-1.8 hrs", "1.8-2 hrs", 
                            "2-2.5 hrs", ">2.5 hrs"))
bins <- c(0, 75, 110, 120, 150, 7140)
pal <- colorBin("YlOrRd", domain = fireprop$TOTAL_INCIDENT_DURATION, bins = bins)
color <- pal(fireprop$TOTAL_INCIDENT_DURATION)

# icon
firehouseIcons <- icons(
  iconUrl = "http://icons.iconarchive.com/icons/icons8/windows-8/48/Maps-Marker-icon.png",
  iconWidth = 15, iconHeight = 20
)

# popup content
# popups

content <- paste("What:",fireprop$INCIDENT_TYPE_DESC,"<br/>",
                 "When:",fireprop$INCIDENT_DATE_TIME,"<br/>",
                 "Where:",fireprop$PROPERTY_USE_DESC,"<br/>", # added where
                 "Location:",fireprop$FIRE_SPREAD_DESC,"<br/>",
                 "Length of Incident:", fireprop$brks, "<br/>",
                 "Number of Units On Scene:",fireprop$UNITS_ONSCENE,"<br/>",
                 "Fire Dectector Status:", fireprop$DETECTOR_PRESENCE_DESC, "<br/>")
 
# new content for firehouse popups
loc_content <- paste("Facility Name:",locs$FacilityName,"<br/>",
                     "Borough:", locs$Borough, "<br/>")


                 
# map
leaflet(options = leafletOptions(minZoom = 8)) %>% 
  setView(lat = 40.730610, lng = -73.935242, zoom = 10) %>%
  addTiles() %>% 
  addProviderTiles("CartoDB.Positron", group = "New York City") %>% 
  addCircleMarkers(data = fireprop, 
                   lat = ~Latitude, lng = ~Longitude, 
                   color = color,
                   opacity = 0.9,
                   fillOpacity = 0.8,
                   weight = 1,
                   radius = ~fireprop$UNITS_ONSCENE/2, # make radius smaller
                   popup = content, 
                   group = "Incidents") %>% 
  addMarkers(data = locs, 
             lat = ~Latitude, lng = ~Longitude, 
             popup = loc_content,
             icon = firehouseIcons,
             group = "Firehouses") %>% 
   addLegend("bottomright", 
            colors = c("#FFFFB2", "#FECC5C", "#FD8D3C", "#F03B20", "#BD0026"), 
            values = ~fireprop$brks, 
            title = "Length of Incident", 
            labels = c("10-75 min", "1.25-1.8 hrs", "1.8-2 hrs", 
                            "2-2.5 hrs", ">2.5 hrs")) %>% 
  addLayersControl(
    baseGroups = "New York City",
    overlayGroups = c("Incidents", "Firehouses"))


```


### 4. Distance from Firehouse and Response Time

I then investigated whether the distance of the incident from the nearest firehouse varied across the city. For all incident locations, I identified the nearest firehouse and calculated the distance between the firehouse and the incident location. 

#### a. Scatterplots 

```{r}
#### calculate distances 
library(geosphere)
library(ggmap)
library(ggthemes)

# firehouse locations lat, lon
fh_loc <- locs[ , 3:4]
colnames(fh_loc) <- c("lat", "lon")
fh_loc <- na.omit(fh_loc)

# incident locations lat, lon
inc_loc <- fire[ , c(25:26,1)]
colnames(inc_loc) <- c("lat", "lon", "IM_INCIDENT_KEY")
inc_loc <- na.omit(inc_loc)

# calculating distance between incident and nearest firehouse
# break up into chunks of 213 due to issues with dfs being different sizes
set1 <- inc_loc[1:213, ]
set2 <- inc_loc[214:426, ]
set3 <- inc_loc[427:639, ]
set4 <- inc_loc[640:852, ]
set5 <- inc_loc[853:1065, ]
set6 <- inc_loc[1066:1278, ]
set7 <- inc_loc[1279:1491, ]
set8 <- inc_loc[1492:1704, ]
set9 <- inc_loc[1705:1917, ]
set10 <- inc_loc[1918:2130, ]
set11 <- rbind(inc_loc[2131:2319, ], inc_loc[1:24,])

sets <- list(set1, set2, set3, set4, set5, set6, set7, set8, set9, set10, set11)
# firehouses_sp <- SpatialPoints(fh_loc)
# df <- NULL
# for (i in sets) {
#   incidents_sp <- SpatialPoints(i) 
#   i$nearest_fh_ID <- apply(gDistance(incidents_sp, firehouses_sp, byid = TRUE), 
#                            1, which.min)
#   i$distance <- round(apply(gDistance(incidents_sp, firehouses_sp, byid = TRUE), 
#                       1, min), 4)
#   df<-rbind(df,i)
# }

df <- NULL
for (i in sets) {
  incidents <- i
  i$nearest_fh_ID <- which.min(distGeo(incidents, fh_loc))
  i$distance <- round((min(distGeo(incidents, fh_loc)) * 0.000621371), 2) # convert to miles
  df<-rbind(df,i)
}

firehouse_incidents <- df[-(2319:2342),] # rid of dummy data
rownames(firehouse_incidents) <- 1:nrow(firehouse_incidents) # reset index

# time until firetruck arrived
full_fire <- merge(x = firehouse_incidents, y = fireprop, by = "IM_INCIDENT_KEY",  all = TRUE)
full_fire$INCIDENT_DATE_TIME <- strptime(full_fire$INCIDENT_DATE_TIME, "%m/%d/%Y %I:%M:%S %p")
full_fire$ARRIVAL_DATE_TIME <- strptime(full_fire$ARRIVAL_DATE_TIME, "%m/%d/%Y %I:%M:%S %p")

full_fire$diff_time <- as.numeric(difftime(full_fire$ARRIVAL_DATE_TIME, full_fire$INCIDENT_DATE_TIME, tz = "EST", units = "mins"))
```

```{r}
##### Scatterplots
#colors = colorRampPalette(brewer.pal(9, "YlOrRd"))(colorCount)
ggplot(full_fire, aes(x = reorder(PROPERTY_USE_DESC, diff_time), y = diff_time)) +
  geom_point(aes(color = diff_time)) + 
  scale_color_gradientn(colors = rev(brewer.pal(3,"YlOrRd"))) +
  #scale_color_manual(values=c("#FFFFCC", "#FFF3B0", "#FEE896", "#FEDC7C", "#FEC662", "#FDAF4A", "#FD9840", #"#FC7936", "#FC522B", "#EE3222", "#DD151D", "#C50523", "#A50026", "#800026")) +
  labs(y = "Arrival Time (min)", x = '',
       title = "NYC Fire Responders: 2015", 
       subtitle = "Arrival times to incidents by property type") +
  coord_flip() +
  theme_tufte() +
  theme(legend.position = "none")
```
```{r}
full_fire$BOROUGH_DESC <-substring(full_fire$BOROUGH_DESC, 5)
ggplot(full_fire, aes(x = as.factor(BOROUGH_DESC), y = diff_time)) +
  geom_point(aes(color = diff_time)) + 
  scale_color_gradientn(colors = rev(brewer.pal(3,"YlOrRd"))) +
  labs(y = "Arrival Time (min)", x = "",
       title = "NYC Fire Responders: 2015", 
       subtitle = "Arrival times to incidents by NYC borough: Bronx the place to be") +
  coord_flip() +
  theme_tufte() +
  theme(legend.position = "none")
```

It doesn't seem like there's much variation in the time it takes fire trucks to get to the incident, even when looking at how far the nearest station is from the property. 

```{r}
ggplot(full_fire, aes(x = distance, y = diff_time)) +
  geom_point(aes(color = diff_time)) + 
  scale_color_gradientn(colors = rev(brewer.pal(3,"YlOrRd"))) +
  labs(y = "Arrival Time (min)", 
       x = "Distance from Nearest Fire Station (mi)",
       title = "NYC Fire Responders: 2015", 
       subtitle = "Arrival times quick & stable regardless of distance to incident") +
  #coord_flip() +
  theme_tufte() +
  theme(legend.position = "none")
```
```{r}
ggplot(full_fire, aes(x = IM_INCIDENT_KEY, y = diff_time)) +
  geom_point(aes(color = diff_time)) + 
  scale_color_gradientn(colors = rev(brewer.pal(4,"YlOrRd"))) +
  labs(y = "Arrival Time (min)", 
       x = "Fire Incidents",
       title = "NYC Fire Responders: 2015", 
       subtitle = "Arrival times quick & relatively stable") +
  #coord_flip() +
  theme_tufte() +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x = element_blank()) +
  theme(legend.position = "none")
```


#### b. Response Times

Lastly, I created a map visualization of response times to the incidents. I differentiated by distance from the nearest firehouse and colored incidents by the duration of the incident. 

```{r}
# categories of distance to fire station
full_fire$distance_rd <- ifelse(full_fire$distance < .25, 
                                "Less Than Quarter Mile", 
                                full_fire$distance)
full_fire$distance_rd <- ifelse(full_fire$distance >= .25 & full_fire$distance < .50, 
                                "Less Than Half Mile", 
                                full_fire$distance_rd)
full_fire$distance_rd <- ifelse(full_fire$distance  >= .5, 
                                "Over Half Mile", 
                                full_fire$distance_rd)

# breaks in fire length
full_fire$brks <- cut(full_fire$TOTAL_INCIDENT_DURATION, 
                   breaks=c(0, 75, 110, 120, 150, 7140), 
                   labels=c("10-75 min", "1.25-1.8 hrs", "1.8-2 hrs", 
                            "2-2.5 hrs", ">2.5 hrs"))

# map                               
ggmap(incident_map) + 
  geom_point(data = full_fire[!is.na(full_fire$distance_rd), ], 
             aes(x = lon, y = lat, color = brks),
                 alpha = 0.5, 
                 size = 1) +
  scale_color_brewer(name = "Length of Fire:", 
                     type = "seq", 
                     palette = "YlOrRd", 
                     direction = 1, 
                     guide = "legend") +
  labs(x = '', y = '') +
  theme_tufte() +
  theme(axis.text = element_blank()) +
  theme(axis.ticks = element_blank()) +
  facet_wrap(~distance_rd, nrow = 1, drop = TRUE) +
  theme(strip.background = element_rect(fill = "#ffff99")) +
  theme(legend.position = "right")
  
```


